Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2010-08-25 Text/migration.tex
Dick changes 8/24/10
git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@69 3b7f7548-5c66-0410-83ab-3997167635fc

Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that is massive enough to experience {\bf any mass object will experience GIs rewrite} gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases, there are times when the planet's radial motion is slow and random. At other times, when the planet interacts with a strong spiral mode near the mode's corotation radius, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.}
%Scott and Aaron: Verify that everything I say here is correct.
The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code (Boley et al. 2007b) is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ (Boley et al. 2007a) and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a resolution {\bf starting resolution is different} ($\varpi$,$\phi$,z) = (512,512,64). The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in Boley et al. (2007b), where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in Michael \& Durisen (2010), and, overall, the planet integration is done using methods similar to those in {\bf integration is done with a verlet integrator as in hut95, yoshida90, channell90, tuckerman92. indirect terms are treated as in nelson} Nelson et al. (2000). The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in Boley et al. (2006, 2007b), except that we correct the D'Alessio et al. (2001) gas mean molecular weight to $\mu$ = 2.39.

>>>>

Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet interacts with a strong spiral mode, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.}
%Scott and Aaron: Verify that everything I say here is correct.
The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code (Boley et al. 2007b) is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ (Boley et al. 2007a) and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a grid size ($\varpi$,$\phi$,z) = (512,512,64). 
{\bf See note in Latex file.} 
%Scott: The resolution stays the same. The grid size varies, but this is unimportant because initially the outer 256 radial zones are empty. It is simplest to just give the grid size. So I did change "resolution" to grid size here.
The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in Boley et al. (2007b), where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in Michael \& Durisen (2010). The planet integration is done with a Verlet integrator (e.g., Hut 1995), and the indirect potential terms are treated as in Nelson et al. (2000). The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in Boley et al. (2006, 2007b), except that we correct the D'Alessio et al. (2001) gas mean molecular weight to $\mu$ = 2.39.
